\documentclass[11pt,]{article}
\usepackage[margin=1in]{geometry}
\newcommand*{\authorfont}{\fontfamily{phv}\selectfont}
\usepackage[]{mathpazo}
\usepackage{abstract}
\renewcommand{\abstractname}{}    % clear the title
\renewcommand{\absnamepos}{empty} % originally center
\newcommand{\blankline}{\quad\pagebreak[2]}

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}} 
\usepackage{longtable,booktabs}

\usepackage{parskip}
\usepackage{titlesec}
\titlespacing\section{0pt}{12pt plus 4pt minus 2pt}{6pt plus 2pt minus 2pt}
\titlespacing\subsection{0pt}{12pt plus 4pt minus 2pt}{6pt plus 2pt minus 2pt}

\titleformat*{\subsubsection}{\normalsize\itshape}

\usepackage{titling}
\setlength{\droptitle}{-.25cm}

%\setlength{\parindent}{0pt}
%\setlength{\parskip}{6pt plus 2pt minus 1pt}
%\setlength{\emergencystretch}{3em}  % prevent overfull lines 

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}

\usepackage{fancyhdr}
\pagestyle{fancy}
\usepackage{lastpage}
\renewcommand{\headrulewidth}{0.3pt}
\renewcommand{\footrulewidth}{0.0pt} 
\lhead{}
\chead{}
\rhead{\footnotesize EC 320: Introduction to Econometrics -- Spring
2022}
\lfoot{}
\cfoot{\small \thepage/\pageref*{LastPage}}
\rfoot{}

\fancypagestyle{firststyle}
{
\renewcommand{\headrulewidth}{0pt}%
   \fancyhf{}
   \fancyfoot[C]{\small \thepage/\pageref*{LastPage}}
}

%\def\labelitemi{--}
%\usepackage{enumitem}
%\setitemize[0]{leftmargin=25pt}
%\setenumerate[0]{leftmargin=25pt}




\makeatletter
\@ifpackageloaded{hyperref}{}{%
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
}
\@ifpackageloaded{color}{
    \PassOptionsToPackage{usenames,dvipsnames}{color}
}{%
    \usepackage[usenames,dvipsnames]{color}
}
\makeatother
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={ ()},
             pdfkeywords = {},  
            pdftitle={EC 320: Introduction to Econometrics},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls


\setcounter{secnumdepth}{0}





\usepackage{setspace}

\title{EC 320: Introduction to Econometrics}
\author{Emmett Saulnier}
\date{Spring 2022}


\begin{document}  

		\maketitle
		
	
		\thispagestyle{firststyle}

%	\thispagestyle{empty}


	\noindent \begin{tabular*}{\textwidth}{ @{\extracolsep{\fill}} lr @{\extracolsep{\fill}}}


E-mail: \texttt{\href{mailto:emmetts@uoregon.edu}{\nolinkurl{emmetts@uoregon.edu}}} & Web: \href{http://emmett.run}{\tt emmett.run}\\
Office Hours: WF 11:00 a.m. - 12:00 p.m.  &  Class Hours: XXX\\
Office: 823 Prince Lucien Campbell  & Class Room: XXX\\
	&  \\
	\hline
	\end{tabular*}
	
\vspace{2mm}
	


\hypertarget{course-description}{%
\section{Course Description}\label{course-description}}

This is the first course in the introductory econometrics sequence.
Econometrics is the science, and art, of using evidence - data - to test
economic theories, predict future events, or make recommendations for
social policies. The primary goals of this course are to provide
students with the knowledge and tools to conduct empirical research and
to critically evaluate statistical analyses by researchers and in the
media. A secondary goal is to show students how prevalent randomness and
chance are in the world, and to teach useful ways to interpret events in
light of this uncertainty.

\hypertarget{textbook}{%
\subsection{Textbook}\label{textbook}}

Dougherty, Christopher, Introduction to Econometrics, Oxford: University
Press (I will assign homework from the 5th edition).

\hypertarget{prerequisites}{%
\subsection{Prerequisites}\label{prerequisites}}

Math 242 (Calculus) and Math 243 (Probability and Statistics)

\hypertarget{grading-policy}{%
\section{Grading Policy}\label{grading-policy}}

Final grades will be based on 8 homework assignments, one midterm, a
final, and possibly quizzes. Any absences must be excused prior to the
time that assignments are due. The final is given only at its stated
time. Your final course grade will be determined as follows:

\begin{itemize}
\item
  \textbf{20\%} of your grade will be determined by homework and
  quizzes.
\item
  \textbf{40\%} of your grade will be determined by a midterm exam.
\item
  \textbf{40\%} of your grade will be determined by a final exam.
\end{itemize}

\hypertarget{homework}{%
\subsection{Homework}\label{homework}}

There will be eight homework assignments in this class. Of these, your
best 7 scores will count toward your final grade. Each assignment will
be due at the beginning of class on the day it is assigned. Late
homework will be accepted until a solution key has been posted.

\hypertarget{quizzes}{%
\subsection{Quizzes}\label{quizzes}}

I reserve the right to give unannounced in class quizzes. If given, they
will be comprised of between five and ten multiple choice questions
designed to show you are keeping up with the assigned materials. Those
not attending class on days where a quiz is given will receive a score
of 0.

\hypertarget{tests}{%
\subsection{Tests}\label{tests}}

There will be one midterm and one final in this class. In the case of a
missed midterm due to unanticipated emergency situations, the student
will be allowed to put the weight of the missed exam on the final exam,
provided notification is received as soon as possible and there is
verification of the emergency. Do not take this class is you already
know you cannot make one of the scheduled exams.

\hypertarget{labs}{%
\subsection{Labs}\label{labs}}

There is a lab associated with this class. Those signed up for the lab
will have use of the computer lab during specific time periods. The lab
will consist of instruction (topics, homework review, grading, etc.) and
computing tips necessary to complete homework assignments. Throughout
the term, a free and open source statistical programming language called
\texttt{R} will be used. You can download the language from
\href{https://cran.r-project.org/}{here}. I also recommend using
\href{https://www.rstudio.com/products/rstudio/download/}{R-Studio} as
your development interface. It is available in the SSIL lab (McKenzie
Hall 4th Floor). Many of the homework assignments will require the use
of \texttt{R}. For more information, contact the SSIL lab. The SSIL lab
also has a wide variety of other programs available.

\hypertarget{academic-dishonesty-policy}{%
\subsection{Academic Dishonesty
Policy}\label{academic-dishonesty-policy}}

Academic dishonesty (from plagiarizing work to cheating on exams) will
not be tolerated. Please acquaint yourself with the Student Conduct
Code, which is published in the Schedule of Classes each term.

\hypertarget{students-with-disabilities}{%
\subsection{Students with
Disabilities}\label{students-with-disabilities}}

If you have a documented disability and anticipate needing
accommodations in this course, please make arrangements with me during
the first week of the term. Please request that the counselor for
students with disabilities (164 Oregon Hall) send me a letter verifying
your disability.

\newpage

\hypertarget{class-schedule}{%
\section{Class Schedule}\label{class-schedule}}

The following is a tentative schedule of lectures. The schedule should
be viewed as tentative and may be subject to change throughout the term.

\hypertarget{week-01-0328---0401-introduction-and-review}{%
\subsection{Week 01, 03/28 - 04/01: Introduction and
Review}\label{week-01-0328---0401-introduction-and-review}}

\hypertarget{week-02-0404---0408-review}{%
\subsection{Week 02, 04/04 - 04/08:
Review}\label{week-02-0404---0408-review}}

\hypertarget{week-03-0411---0415-simple-regression-analysis}{%
\subsection{Week 03, 04/11 - 04/15: Simple Regression
Analysis}\label{week-03-0411---0415-simple-regression-analysis}}

\hypertarget{week-04-0418---0422-properties-of-coefficients}{%
\subsection{Week 04, 04/18 - 04/22: Properties of
Coefficients}\label{week-04-0418---0422-properties-of-coefficients}}

\hypertarget{week-05-0425---0429-hypothesis-testing}{%
\subsection{Week 05, 04/25 - 04/29: Hypothesis
Testing}\label{week-05-0425---0429-hypothesis-testing}}

\hypertarget{week-06-0502---0506-multiple-regression-analysis}{%
\subsection{Week 06, 05/02 - 05/06: Multiple Regression
Analysis}\label{week-06-0502---0506-multiple-regression-analysis}}

\hypertarget{week-07-0509---0513-multicollinearity}{%
\subsection{Week 07, 05/09 - 05/13:
Multicollinearity}\label{week-07-0509---0513-multicollinearity}}

\hypertarget{week-08-0516---0520-nonlinearity-and-transformation}{%
\subsection{Week 08, 05/16 - 05/20: Nonlinearity and
Transformation}\label{week-08-0516---0520-nonlinearity-and-transformation}}

\hypertarget{week-09-0523---0527-dummy-variables}{%
\subsection{Week 09, 05/23 - 05/27: Dummy
Variables}\label{week-09-0523---0527-dummy-variables}}

\hypertarget{week-10-0530---0603-model-specification}{%
\subsection{Week 10, 05/30 - 06/03: Model
Specification}\label{week-10-0530---0603-model-specification}}




\end{document}

\makeatletter
\def\@maketitle{%
  \newpage
%  \null
%  \vskip 2em%
%  \begin{center}%
  \let \footnote \thanks
    {\fontsize{18}{20}\selectfont\raggedright  \setlength{\parindent}{0pt} \@title \par}%
}
%\fi
\makeatother