---
title: "**Working with Data**"
author: |
        | **Modifications:** Giorgi Nikolaishvili
        | **Original Authors:** Undetermined Group of Past PhD Students
        | EC 320: Introduction to Econometrics
        | University of Oregon
date: "Spring 2022"
output: 
  html_document:
    theme: flatly  
    includes:
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: true
      
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, warning = F, message = F, cache = T)
```

## Recap 

What we learned last week:

- Math operators in R
- Variables (objects) in R
- Functions in R
- Variable classes in R
- Loading packages (using `pacman`)
- Data frames
- Basic plotting (using `ggplot2`)

Some notes on the computational assignment and other more general notes:

- You should never hard-code / manually copy any values from a dataset!

```{r}
x <- mean(mtcars$mpg)
x
```

- `.Rmd` files don't knit when your code has a line with `install.packages()` in which you are installing a package that already exists on your machine. 
Generally, you should only use `install.packages()` in the console -- after you've installed a package once, it will exist on your computer forever.
Many of you put the `install.packages("pacman")` line in your `.Rmd`s and ran into knitting issues -- after the initial installation of `pacman` you only need to use `library(pacman)` to load the `pacman` package, and then simply use the `p_load` function to install/load other packages you need.

- Get started with your computational assignments early! 
There is no guarantee that I will be able to help you out if you run into issues on the day of the deadline.

- If you have questions, you should: **(1)** come to my office hours; **(2)** if you can't do (1) then you should send me an email in which you state a specific question or problem you are running into; **(3)** if you don't have a specific question/problem and instead are simply confused/lost/unsure, then refer to (1) or email me to schedule a one-on-one meeting in which I can set you on the right track. 

- Please don't email me an attachment of all of your work with the question ``Am I doing everything right?'' :~)
I can't answer a general question like that for a variety of practical and ethical reasons.

- **Attend the labs or watch the full recording. Not every piece of necessary information will be given in these lab note documents -- I might mention and demonstrate crucial concepts on the fly.**

## Writing and Reading `.csv` Files

The `mtcars` dataset comes pre-loaded with R for us to play around with.

Suppose I want to save `mtcars` as a `.csv` file on my computer.

I would use the `write.csv` function to do so. 

The first input to this function is the data frame object in question.
The second input is the path for the `.csv` file that is to be written (including the name of the file at the very end). 

```{r}
# Save `mtcars` dataframe as `.csv` file 
write.csv(mtcars, "C:/Users/gioni/Box/uo_phd_y3/spring_ec320/EC320s22/labs/02-working-with-data/data.csv")
```

In the above chunk, I've saved `mtcars` as a `.csv` file called `data.csv` in a specific directory (folder) on my computer.

Now, suppose I would like to import the data in `data.csv` and store it in a dataframe called `df` in my current instance of R. 

I would use the `read.csv` function to do so. 

```{r}
# Import `data.csv` and store as data frame
df <- read.csv(file = "C:/Users/gioni/Box/uo_phd_y3/spring_ec320/EC320s22/labs/02-working-with-data/data.csv")
```

Let's preview the new data frame `df` to convince ourselves that the above code did its job:

```{r}
df
```

The `read.csv` function automatically saves the dataset as a data frame:

```{r}
class(df)
```

Alternatively, we could convert the data frame object `df` to a "tibble":

```{r}
df <- as.tbl(df)
df
```

Tibble is an extremely similar alternative object class to data frame that is provided by the `tidyverse` package. 
We will not go into details on tibble vs. data frame here, but the above was worth showing since you are asked to convert a data frame object to a tibble in your next computational assignment.
For now, you can think of tibbles and data frames as being interchangeable object classes.

## Working Directory

If you have a large hierarchy of directories (folders) on your computer like I do, it may get cumbersome to keep typing out the full paths of datasets you want to import.

For this reason, it is useful give R directions on how to reach your desired dataset relative to your present **working directory** -- this is the directory (folder) that R treats as an anchor (where R thinks you are currently "working"). 

Find your current directory: `getwd()`.
"wd" stands for "working directory". 

```{r}
getwd()
```

Show files in the current directory: `dir()`

```{r}
dir() 
```
Change directory: `setwd()`

```{r}
setwd("C:/Users/gioni/Box/uo_phd_y3/spring_ec320/EC320s22/labs/02-working-with-data/newfolder")
```

Notice that directories are characters (surrounded by `""`). 

RStudio will help you complete file paths when you hit tab while typing in the console.

You can save an object with the directory location and then return to that directory using `setwd()`. 

Example:

```{r}
my_dir <- "C:/Users/gioni/Box/uo_phd_y3/spring_ec320/EC320s22/labs/02-working-with-data"
setwd(my_dir)
```

Now let's try importing the `data.csv` file relative to our working directory:

```{r}
df <- read.csv(file = "data.csv")
```

What if we wanted to write the `data.csv` file in the `newfolder` subdirectory instead?

```{r}
write.csv(mtcars, "newfolder/data.csv")
```

To import the above dataset:

```{r}
df <- read.csv(file = "newfolder/data.csv")
```

The above paths to the `.csv` file relative to the current working directory are called **relative paths**. 

## Tidy Data

Data can be **messy**, or it can be **tidy**. 

**Tidy data** has the following properties:

- Every columns is a variable.
- Every row is an observation.
- Every cell is a single value.

**Messy data** is any other arrangement of data.
It can come in many forms, and usually needs to be thoroughly "cleaned" before it can be analyzed.
Tidy data makes it easy for an analyst or a computer to extract needed variables because it provides a standard way of structuring a dataset.

### Example: `gapminder`

`gapminder` is a dataset that lives in the `gapminder` R package and contains data on countries' total population, GDP per capita, and life expectancy over time.
The dataset contains the following variables:

- `gdpPercap`: GDP per capita of a country in a given year.
- `pop`: Population of a country in a given year.
- `lifeExp`: Life expectancy in a country in a given year.
- `year`: The year in which an observation was taken.
- `continent`: The continent to which a country associated with a given observation belongs.
- `country`: The country with which a given observation is associated.

Let's load the `tidyverse` and `gapminder` packages to start working with this data.

```{r}
library(pacman)
p_load(tidyverse, gapminder)
```

Print the dataset:

```{r}
gapminder
```

Print the first few observations of the dataset using the `head()` function:

```{r}
head(gapminder)
```

Print the last few observations of the dataset using the `tail()` function:

```{r}
tail(gapminder)
```

Print basic summary statistics on each of the variables contained by the dataset using the `summary()` function:

```{r}
summary(gapminder)
```

## Selecting Subsamples 

Suppose we do not care about country population data.

Let's the `country`, `year`, and `pop` variables and store it all together in a new tibble object called `df_countrypop`. 

The `tidyverse` package provides us a handy function called `select` that allows us to grab a select set of variables from a data frame. 

```{r}
df_countrypop <- gapminder %>% select(country, year, pop)
df_countrypop
```

The pipe operator, written as `%>%`, takes the output of one function and passes it into another function as an argument. 
This allows us to link a sequence of analysis steps.

In the above code chunk, the pipe operator automatically gives `select()` the `gapminder` tibble as an input, so that `select()` knows that `country`, `year`, and `pop` are columns that belong to `gapminder`. 

The pipe operator is extremely useful! 
Start playing around with it -- you will rely on it heavily this quarter.

Alternatively (note that this is bad practice) you could have written

```{r}
# DO NOT do the following in your code!!!
df_countrypop <- select(gapminder, country, year, pop) # This is bad.
df_countrypop 
```

Now, suppose we only care about population data on the United States.
Let's take the `df_countrypop` tibble we've just created and filter for US-related data only using the `filter()` function. 
Once again, we should use the pipe operator.

```{r}
df_usapop <- df_countrypop %>% filter(country == "United States")
df_usapop
```

The above sequence of steps is equivalent to having done the following in one go:

```{r}
df_usapop <- gapminder %>%
              select(country, year, pop) %>%
              filter(country == "United States")
```

Note the use of `==` inside of `filter` in the above code chunk.
A single equal sign `=` acts as an assignment operator, while the double equal sign `==` is a **relational** operator that yields "TRUE" if two objects are equal and "FALSE" otherwise.

The `filter(country == "United States")` command basically skims through all observations inside of the `gapminder %>% select(country, year, pop)` dataset and for each observation says "if the value of the `country` variable in this observation is equal to "United States", then keep this observation -- if not, then discard this observation".
Or equivalenty "if `country == "United States"` is true for this observation, then keep it -- discard otherwise." 

Let's quickly plot the data we've gathered in `df_usapop`:

```{r}
ggplot(df_usapop, aes(x = year, y = pop)) +
  geom_line()
```

## Transforming Data

We can easily create a new column of data and add it to `df_usapop` by transforming the existing columns.

For example, suppose we want to express the total population values `pop` in millions and store it in a new column called `popmil`: 

```{r}
df_usapop <- df_usapop %>%
              mutate(popmil = pop/1000000)
df_usapop
```

Let's visual `popmil` now as a timeplot.

```{r}
ggplot(df_usapop, aes(x = year, y = popmil)) +
  geom_line()
```


## Ordering Data

Now let's focus on the life expectancy data of all countries in the year 1992.
We can easily gather the above data in a new object called `df_lifeExp92` using some of the tools we've learned so far:

```{r}
df_lifeExp92 <- gapminder %>% 
                select(country, year, lifeExp) %>%
                filter(year == 1992)
df_lifeExp92
```

We may use the `arrange()` function to order all of the observations in `df_lifeExp92` in increasing order with respect to the values in `lifeExp`.

```{r}
df_lifeExp92 <- df_lifeExp92 %>%
                arrange(lifeExp)
df_lifeExp92
```

Similarly, we may us ethe `desc()` function inside of `arrange()` to order the observations in decreasing order with respect to `lifeExp`. 

```{r}
df_lifeExp92 <- df_lifeExp92 %>%
                arrange(desc(lifeExp))
df_lifeExp92
```
