---
title: "**Grouping, Summarizing, and Plotting Data**"
author: |
        | Giorgi Nikolaishvili
        | EC 320: Introduction to Econometrics
        | University of Oregon
date: "Spring 2022"
output: 
  html_document:
    theme: flatly  
    includes:
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: true
      
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, warning = F, message = F, cache = T)
```

## Recap 

What we learned last week:

- Reading and writing `.csv` files
- Absolute vs. relative file paths
- Tidy data
- `select()` -- select variables
- `filter()` -- select observations
- `mutate()` -- create new variables
- `arrange()` -- order observations
- `%>%` -- piping

## Loading Packages (Review)

Suppose we want to load the `tidyverse` and `gapminder` packages.

An **incorrect** approach:

```{r eval=FALSE}
library(tidyverse)
library(gapminder)
```

Another **incorrect** approach:

```{r eval=FALSE}
library(pacman)
p_load(tidyverse)
p_load(gapminder) 
```

The **correct** approach:

```{r}
library(pacman) 
p_load(tidyverse, gapminder)
```

## Pipe Operator `%>%` (Review) 

Recall that the pipe operator, `%>%`, passes an object to a function as an input.

An **incorrect** approach to applying multiple functions in succession:

```{r eval=FALSE}
df <- select(mtcars, mpg, wt) # select `mpg` and `wt` columns from mtcars
df <- filter(df, mpg >= 20)   # select observations with `mpg` > or = 20
df <- arrange(df, mpg)        # order observations with respect to `mpg`
```

Another **incorrect** approach:

```{r eval=FALSE}
df <- arrange(filter(select(mtcars, mpg, wt), mpg >= 20), mpg)
```

The **correct** approach using the pipe operator:

```{r}
df <- mtcars %>%  
        select(mpg, wt) %>%     # `select()` receives `mtcars`   
        filter(mpg >= 20) %>%   # `filter()` receives `mtcars %>% select()`
        arrange(mpg)            # `arrange()` receives `mtcars %>% select() %>% filter()`
```

## Grouping and Summarizing

As econometricians, we sometimes want to aggregate and produce summary statistics about our data.

Recall the `gapminder` dataset:

```{r}
head(gapminder) 
```

- What is the average/smallest/greatest life expectancy across all countries over the entire sample period?

- What is the total GDP of the world in the year 1997?

- What is the average/smallest/greatest life expectancy over the entire sample period in each continent?

- What is the average/smallest/greatest life expectancy in each year in each continent?

- What is the total GDP in each year in each continent?

The above questions can be answered with the aid of the `summarize()` and `group_by` functions, available through `tidyverse`. 

### `summarize()` 

The `summarize()` function takes a dataset as an input, along with instructions on how to summarize the given data and store the indicated summary into a new tibble as a variable (or multiple variables).

Simple example:

```{r}
gapminder %>%
  summarize(avgLifeExp = mean(lifeExp))
```

In the above code chunk, we've given `summarize()` a data input of `gapminder`, and then told it to compute the average of the `lifeExp` column using `mean()`, and store the output into a variable called `avgLifeExp`. 

Now let's move on to other questions, and combine `summarize()` with other familiar functions (`filter()`, `select()`, `mutate()`, `arrange()`). 

What was the average life expectancy in 1997?

```{r}
gapminder %>%
  filter(year == 1997) %>%
  summarize(avgLifeExp = mean(lifeExp))
```

What was the total GDP of the world in 1997 (in 1000s of USD)?

```{r}
gapminder %>%
  filter(year == 1997) %>%
  mutate(gdp = pop * gdpPercap) %>%
  summarize(totGDP = sum(gdp)/(10^9)) 
```

We can answer both of the above questions inside of a single `summarize()` command:

```{r}
gapminder %>%
  filter(year == 1997) %>%
  mutate(gdp = pop * gdpPercap) %>%
  summarize(avgLifeExp = mean(lifeExp),
            totGDP = sum(gdp)/(10^9)) 
```

You can use the following functions inside of `summarize()`:

- `mean()`
- `sum()`
- `median()`
- `min()`
- `max()` 

### `group_by()`

The `group_by()` function allows us to group the observations in a dataset into different categories.

For example, in the `gapminder` dataset, we know that each observation is associated with a specific country, located in a specific continent, at a specific point in time.

Therefore, we could partition our data by 

1. country (`group_by(country)`)
2. continent (`group_by(continent)`)
3. time (`group_by(year)`)
4. country and time (`group_by(country, year)`) 
5. continent and time (`group_by(continent, year)`)

and produce summary statistics for each partition of the data.

Let's see `group_by()` in action by answering some questions about the `gapminder` data.

What is the total GDP of each continent in the year 1997?

```{r}
gapminder %>%
  filter(year == 1997) %>%
  group_by(continent) %>%                 # 
  mutate(gdp = pop * gdpPercap) %>%
  summarize(avgLifeExp = mean(lifeExp),
            totGDP = sum(gdp)/(10^9)) 
```

What is the total world GDP in each year of the sample?

```{r}
gapminder %>%
  group_by(year) %>%
  mutate(gdp = pop * gdpPercap) %>%
  summarize(avgLifeExp = mean(lifeExp),
            totGDP = sum(gdp)/(10^9)) 
```

What is the total GDP of each continent in each year of the sample?

```{r}
gapminder %>%
  filter(year == 1997) %>%
  group_by(continent, year) %>%                 # 
  mutate(gdp = pop * gdpPercap) %>%
  summarize(avgLifeExp = mean(lifeExp),
            totGDP = sum(gdp)/(10^9)) 
```

## Plotting using `ggplot2`

`ggplot2` is a package that lives inside `tidyverse` (`tidyverse` is actually a package that contains a group of packages, each of which contains a group of functions).

`ggplot2` allows us to neatly visualize data! 

Many academic researchers and data scientists exclusively use `ggplot2` to make graphs.

Today we will learn the basics of `ggplot2` and work through some examples. 

Click [here](https://ggplot2-book.org/index.html) for a wonderful free text on using ggplot.

### Scatter Plots

Suppose I want to focus on `gapminder` data of the year 1997.

```{r}
data_97_df <- gapminder %>%
            filter(year == 1997) 
```

I suspect that GDP per capita and life expectancy are likely to be positive correlated -- in other words, higher life expectancy is likely to be associated with higher GDP per capita, and vice versa.

Let's test this hypothesis non-rigorously by plotting the data in a scatterplot:

```{r}
ggplot(data_97_df, aes(x = gdpPercap, y = lifeExp)) + # Contents of the graph
  geom_point()                                        # Type of graph
```

Seems like my hypothesis was correct (for the year 1997, at least).

I know that each data point in the above scatter plot is associated with a country, but which continent is each point associated with?

I would like for my graph to show me the continent associated with each point by assigning a color to each continent, and then coloring its corresponding data points accordingly.

```{r}
ggplot(data_97_df, aes(x = gdpPercap, 
                       y = lifeExp, 
                       color = continent)) +
  geom_point()
```

Great! But I'm still not quite happy with the above graph. 

The points look a little too small.

We can modify the properties of the plot inside of `geom_point`. 

To change point size, we can simply set `size = x`, where `x` is some numeric value.

```{r}
ggplot(data_97_df, aes(x = gdpPercap, 
                       y = lifeExp, 
                       color = continent)) +
  geom_point(size = 2.5)
```

Let's also add a title using `ggtitle()`, as well as axis and legend labels using `labs()`. 

```{r}
ggplot(data_97_df, aes(x = gdpPercap/1000,
                       y = lifeExp, 
                       color = continent)) +
  geom_point(size = 2.5) +
  ggtitle("Life Expectancy vs. Output") +
  labs(x = "GDP per capita (USD, 1000s)",
       y = "Life Expectancy (Years)",
       color = "Continent")
```

Lastly, we can manually set the intervals of the x and y axes using `xlim()` and `ylim()`, respectively.

```{r}
  ggplot(data_97_df, aes(x = gdpPercap/1000, 
                         y = lifeExp, 
                         color = continent)) +
    geom_point(size = 2.5) +
    ggtitle("Life Expectancy vs. Output") +
    labs(x = "GDP per capita (USD, 1000s)",
         y = "Life Expectancy (Years)",
         color = "Continent") +
    ylim(0,100) +
    xlim(0,50) 
```

### Line Plots 

Now let's move on to line plots.

Line plots are often used to visualize the changes in a variable over time.

Suppose I am interested in the average GDP per capita of each continent for each year in my `gapminder` sample -- I can use `group_by()` and `summarize()` to obtain this type of aggregated data from the raw country-level data.

```{r}
continent_allyears_df <- gapminder %>%
                          group_by(year, continent) %>%
                          summarize(avgGDPPercap = mean(gdpPercap))

continent_allyears_df
```

I can now create a line graph of my continent-level data, in which the x-axis represents time and the y-axis represents the average GDP per capita in 1000s of USD.

```{r}
ggplot(continent_allyears_df, 
        aes(x = year, 
            y = avgGDPPercap/1000, 
            color = continent)) + 
    geom_line()
```

We can improve the plot by making the lines thicker and adding title, axis, and legend labels. 

```{r}
ggplot(continent_allyears_df, 
        aes(x = year, 
            y = avgGDPPercap/1000, 
            color = continent)) + 
    geom_line(size = 1) +
    ggtitle("Continent GDP per capita Over Time") +
    labs(x = "Year",
         y = "Average GDP per capita (USD, 1000s)",
         color = "Continent")
```

The may further improve the plot by getting rid of excess empty intervals on the x-axis.

```{r}
ggplot(continent_allyears_df, 
        aes(x = year, y = avgGDPPercap/1000, color = continent)) + 
    geom_line(size = 1) +
    ggtitle("Continent GDP per capita Over Time") +
    labs(x = "Year",
         y = "Average GDP per capita (USD, 1000s)",
         color = "Continent") +
    scale_x_continuous(expand = c(0,0))
```

### Histograms

Sometimes we want to check the distributions of variables within our sample -- to do so we may use histograms.

Suppose we're interested in checking the distribution of life expectancy in a subsample of `gapminder` containing observations from only the years 1957 and 2007. 

```{r}
data_57_97_df <- gapminder %>%
                  filter(year %in% c(1957,2007))

data_57_97_df
```

The conditional statement `year %in% c(1957,2007)` checks whether for each given observation `year` equals either 1957 or 2007 -- if `TRUE`, then `filter()` keeps these observations.

We may now plot a simple histogram of `lifeExp` using this newly-constructed subsample of `gapminder`.

```{r}
ggplot(data_57_97_df, aes(lifeExp)) +
    geom_histogram()
```

Add title and axis labels:

```{r}
ggplot(data_57_97_df, aes(lifeExp)) +
    geom_histogram() +
    ggtitle("Life Expectancy Frequency Plot") + 
    labs(x = "Life Expectancy (Years)",
         y = "Count")
```

Intuitively, the distribution of `lifeExp` in this subsample is bimodal.

This is because there are two distinct distributions generating two sets of observations in our dataset -- life expectancy in 1957 for each country is likely to be much lower than in 1997.

Let's group the observations in our histogram by year:

```{r}
ggplot(data_57_97_df, aes(lifeExp, fill = factor(year))) +
    geom_histogram() +
    ggtitle("Life Expectancy Frequency Plot") + 
    labs(x = "Life Expectancy (Years)",
         y = "Count", 
         fill = "Year")
```

Perhaps we would like to make the bins wider:

```{r}
ggplot(data_57_97_df, aes(lifeExp, fill = factor(year))) +
    geom_histogram(binwidth = 5) +
    ggtitle("Life Expectancy Frequency Plot") + 
    labs(x = "Life Expectancy (Years)",
         y = "Count", 
         fill = "Year")
```

Instead of stacking the two groups of data on top of each other, we might instead like to give each group a separate bar per bin:

```{r}
ggplot(data_57_97_df, aes(lifeExp, fill = factor(year))) +
    geom_histogram( binwidth = 5,
                    position = "dodge") +
    ggtitle("Life Expectancy Frequency Plot") + 
    labs(x = "Life Expectancy (Years)",
         y = "Count", 
         fill = "Year")
```